For submission to the Journal of Integrated Pest Management

### 2021-08-12 ofv_fife_2021.Rmd
making adjustments based on reviewer's suggestions

### First report of the *Brevipalpus*-transmitted (Trombidiformes: Tenuipalpidae) *Orchid fleck dichorhavirus* infecting three ornamentals in Florida

@@

We describe the first outbreaks of *Orchid fleck dichorhavirus*, belonging to the orchid-infecting subgroup (OFV-Orc), from three unreported hosts: *Liriope muscari*, cv. 'Gigantea' (Decaisne) Bailey, *Ophiopogon intermedius* Don and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) in Leon and Alachua Counties, FL. Strains of OFV-Orc infect over 50 plant species belonging to the plant families Orchidaceae, Asparagaceae (Nolinoidaea), and infects *Citrus* (Rutaceae) as citrus leprosis disease. The only known vectors of OFV-Orc are flat mites, genus *Brevipalpus* (Trombidiformes: Tenuipalpidae). Florida has various plants in the landscape which *Brevipalpus* spp. feed on, which are susceptible to infection by OFV-Orc.

@@

Two orchid-infecting strains of Orchid fleck virus were detected via combinations of conventional RT-PCR, RT-qPCR, Sanger sequencing and High Throughput Sequencing. Amplicons shared 98% nucleotide identity with OFV-Orc1 and OFV-Orc2 available in NCBI GenBank. Coinfections were seen in each county, but single strains of OFV-Orc were seen in *L. muscari* (Alachua, OFV-Orc2) and *A. elatior* (Leon, OFV-Orc1).

@@

In conclusion, *Orchid fleck dichorhavirus* is present in northern Florida, representing a risk for susceptible plants in the southeastern US.

### Keywords:
False spider mite, flat mite, *Brevipalpus*-transmitted viruses, *Liriope*, Nolinoidaea, *Ophiopogon*, *Aspidistra*, Ruscaceae, Rutaceae, Asparagaceae, orchid, Orchidaceae, pests, ornamental plants, Orchid fleck virus.

\pagebreak
*Orchid fleck dichorhavirus*, commonly referred to as Orchid Fleck Virus (OFV), is the type member for the genus *Dichorhavirus*, family *Rhabdoviridae*. OFV is a bacilliform, nuclear rhabdovirus composed of two segments of single-stranded, negative-sense RNA which infects plants [@Amarasinghe2019; @Dietzgen2014; @Walker2018]. Flat mites from the genus *Brevipalpus* (Trombidiformes: Tenuipalpidae) are the only group known to transmit dichorhaviruses [@Maeda1998], and *Brevipalpus californicus* (Banks) *sensu lato* are the only mites which do so in a persistent propagative manner [@Kondo2003].

@@

Viral infections of suspected leaf samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. All the samples were tested with one step conventional RT-PCR, and were found negative for begomovirus, carlavirus, potyvirus, tospovirus, *Cucumber mosaic virus* and *Tobacco mosaic virus*.

As initial diagnostics were inconclusive, new samples were collected during July and August of 2020 to collect more of these putatively infected plants with ringspot symptoms. The plants collected included *Liriope* spp. and *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea). *A. elatior* was suspected to be infected, due to both its proximity to infected *Liriope* and the presence of unusually chlorotic leaves (`r citef("fig_2")`). Upon collection, the new samples were sent to the Florida Department of Agriculture and Consumer Services (FDACS) for identification.

The FDACS determined eventually that the pathogen was *Orchid fleck dichorhavirus* using previously published primers and methods to conduct RT-PCR and Sanger sequencing [@Kubo2006a; @Kubo2006b; @Kubo2009a; @Kubo2009b; @RamosGonzalez2015]. Orchid subgroup 1, OFV-Orc was identified following the methods described in @Kondo2017. Sequencing demonstrated a shared 98\% nucleotide identity with the orchid strain subgroup, OFV-Orc (isolates So and Br with GenBank Accession numbers: AB244418 and MK522807, respectively) [@Kondo2006; @Kondo2017].

These samples from FDACS were subsequently retested by the USDA-ARS, in conjunction with tests of fresh samples from both Alachua and Leon counties. The USDA used RT-PCR, RT-qPCR, and High Throughput Sequencing (HTS) in sequence to reconfirm the presence of *Orchid fleck dichorhavirus*. RT-PCR and qPCR with Generic R2-Dicho-GF and R2-Dicho-GR primers amplifed ~800 nt of L-gene (RNA2) amplicon [@Roy2020], and OFV-Orc1 and OFV-Orc2 were detected in both *O. intermedius* and *A. elatior* from Leon County.

@@

After the initial identification by FDACS of OFV-Orc strains, mite samples were collected from symptomatic Asparagaceae in Leon County. Most mites collected were Tenuipalpid mites (flat mites or false spider mites), a known pest of ornamental plants, some of which are known to act as vectors for plant viruses [@Childers2003; @Childers2011].

@@

The flat mites collected were initially suspected to belong to *B. californicus* after inspection with phase contrast microscopy. Subsequent observation via DIC microscopy at FDACS agreed with this tentative identification. Unfortunately, the *B. californicus* s.l. species group, *sensu* @Baker1987 is suspected to contain cryptic species [@Childers2011; @Rodrigues2013]. New mite samples were collected from symptomatic liriopogons and *A. elatior* in Leon County and sent to USDA-ARS's Electron and Confocal Microscopy Unit for analysis. Three mite species were recovered and examined under cryo-scanning electron microscopy (Cryo-SEM): *B. californicus* s.l. (`r citef("fig_3")`), *B. obovatus* Donnadieu and *B. confusus* Baker (`r citef("fig_4")`).

@@

OFV consists of two orchid strains (OFV-Orc1 and OFV-Orc2) and two citrus strains (OFV-Cit1 and OFV-Cit2) [@Roy2020; @Beltran-Beltran2020]. The OFV strains detected in Florida are identical in gene order, content, and genome sequence to the orchid strains of OFV infecting citrus in Hawaii, Mexico, Colombia, and South Africa [@Roy2020; @Beltran-Beltran2020]. Both OFV-Orc1 and OFV-Orc2 infect citrus [@Roy2020], but none of the citrus strains have been reported from any orchid species. The *Brevipalpus* mites collected from liriopogons and *A. elatior* in Leon County were abundant on OFV-infected plants very near to citrus trees, some plants even surrounding the trunk. *B. californicus* s. l. has been reported as a pest of citrus [@Childers2003] and are often collected from citrus rinds [@Baker1949; @Baker1987]. The proximity of these mite vectors to citrus raises the question: why these trees are not currently infected with OFV-Orc? It is important to note the uncertainty surrounding the vector for OFV-Orc. There are three mite species which have been recovered from OFV-Orc infected plants: *B. californicus* s.l. (the most likely culprit), *B. obovatus*, and *B. confusus*. Each species has its own unique biology, and all have been implicated with a variety of different hosts. This suggests that the spread of OFV-Orc would be a function of various combinations of several potential factors, including host preferences, vectorial capacity, viral propagation/circulation in the vector, viral acquisition times, and feeding times required for transmission. Some of these factors have been tested: *Brevipalpus* vectors have demonstrated such virus-vector specificity with *Citrus leprosis virus N*--another dichorhavirus which causes citrus leprosis [@Roy2013a]--in studies done by @Ferreira2020 and @GarciaEscamilla2018. In these studies, *Brevipalpus* mites were not able to transmit more than one virus. This could mean that the *B. californicus* which we find on liriopogons and *A. elatior* are not actually the same species as those found on citrus, or at least are not able to transmit OFV to citrus.

@@

In conclusion, detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivated *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and almost eradicated by the mid-1960s [@Childers2003; @Knorr1968a; @Knorr1968b]. An examination of herbarium specimens of Florida citrus found that this historical virus, Citrus leprosis dichorhavirus-N0, is distantly related to the modern strains of OFV [@Hartung2015; @Kitajima2011a; @Roy2020]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the threat of different strains of OFV on citrus, which will be a definite concern to the US multi-billion-dollar citrus industry already impacted by the Huanglongbing disease. *B. californicus*, as well as *B. yothersi*, are both known vectors of dichorhaviruses (OFV) [@GarciaEscamilla2018; @Kondo2003; @Beltran-Beltran2020] and *B. obovatus* is a suspected vector as well [@Childers2003]. All three mite species/complexes are present in Florida [@Childers2003; @Akyazi2017] (`r citef("fig_4")`), and are difficult to identify by non-experts, or without advanced methodologies. DNA barcoding [@Armstrong2005] or a similarly simple and accurate method for identification of these mite complexes is vital to determine which of these species are responsible for transmission of OFV-Orc, and therefore which mite populations need to be monitored or controlled. By doing so, we can determine the risk OFV-Orc represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

@@

### Table 1: List of Asparagaceae (Nolinoidaea) species with verified cases of *Orchid fleck dichorhavirus*, collected from the landscape of northern Florida

|Scientific Name|Common Names|County|Strains|
|:-----------|:--------------|:-----|:------|
|*Liriope muscari* cv. 'Gigantea'\* (Decaisne) Bailey | Lilyturf, Orchardgrass, Monkeygrass|Alachua & Leon|OFV-Orc1 & OFV-Orc2|
|*Ophiopogon intermedius*\*\* Don| Aztec Grass, 'Argenteomarginatus'|Leon|OFV-Orc1 & OFV-Orc2|
|*Aspidistra elatior* Blume| Cast Iron Plant, Bar-room Plant|Leon|OFV-Orc1 & OFV-Orc2|
`r tab_nums(name = "tab_1", caption = "* *Liriope muscari* cv. 'Gigantea' has been traditionally classified as *L. gigantea* Hume by @Broussard2007 and @Fantz2015, although this distinction has been challenged by @Wang2014 and @Masiero2020. * * *O. intermedius* is sometimes misclassified as *Liriope muscari* ‘Variegated Evergreen Giant’ [@Fantz2008b, @Fantz2009] or ‘Grandiflora White’ [@Fantz2009].")`

### Figure captions
`r fig_nums(name = "fig_1", caption = "Variety of symptoms expressed by *Liriope* spp. infected with *Orchid fleck dichorhavirus*: (a) ringspot symptoms on *Liriope muscari* cv. 'Gigantea' (b-c) Details of ringspot symptoms on *L. muscari* cv. 'Gigantea' (d) rust colored spots on *Ophiopogon intermedius*")`

`r fig_nums(name = "fig_2", caption = "Symptoms expressed by *Aspidistra elatior* infected with *Orchid fleck dichorhavirus*: (a) Detail of leaf chlorosis (b) Chlorosis appears similar to sun damage (c-d) Chlorotic ringspot may indicate early symptoms of OFV")`

@@

`r fig_nums(name = "fig_4", caption = "Florida is home to other common pest species of *Brevipalpus*, which are potential vectors of *Orchid fleck dichorhavirus*: (a) *B. phoenicis*, dorsal (b) *B. yothersi*, lateral (c) *B. obovatus*, dorsal.")`

@@

```{r, fig_4, warning=FALSE}
grid::grid.raster(tiff::readTIFF("figure/other_mites.tif"))
```

>>>>

### **Request from Reviewer 4: Revise title to better reflect content of publiction** First report of the *Brevipalpus*-transmitted (Trombidiformes: Tenuipalpidae) orchid fleck virus infecting three ornamentals in Florida

@@

We describe the first detection of orchid fleck virus (OFV), belonging to the orchid-infecting subgroup (OFV-Orc), infecting three unreported hosts: *Liriope muscari*, cv. 'Gigantea' (Decaisne) Bailey, *Ophiopogon intermedius* Don and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) in Leon and Alachua Counties, FL. Strains of OFV-Orc infect over 50 plant species belonging to the plant families Orchidaceae, Asparagaceae (Nolinoidaea), and infects *Citrus* (Rutaceae) as citrus leprosis disease. The only known vectors of OFV-Orc are the flat mites  *Brevipalpus californicus* (Banks) *sensu lato* (Trombidiformes: Tenuipalpidae). Florida has various plants in the landscape which *Brevipalpus* spp. feed on, which are susceptible to infection by OFV-Orc.

@@

Two orchid-infecting strains of OFV were detected via combinations of conventional RT-PCR, RT-qPCR, Sanger sequencing and High Throughput Sequencing. Amplicons shared 98% nucleotide identity with OFV-Orc1 and OFV-Orc2 available in NCBI GenBank. Coinfections were seen in each county, but single strains of OFV-Orc were seen in *L. muscari* (Alachua, OFV-Orc2) and *A. elatior* (Leon, OFV-Orc1).

@@

In conclusion, OFV is present in northern Florida, representing a risk for susceptible plants in the southeastern US.

### Keywords:
False spider mite, flat mite, *Brevipalpus*-transmitted viruses, *Liriope*, Nolinoidaea, *Ophiopogon*, *Aspidistra*, Ruscaceae, Rutaceae, Asparagaceae, orchid, Orchidaceae, pests, ornamental plants, orchid fleck virus.

\pagebreak
Orchid fleck virus (OFV), is the type member for the genus *Dichorhavirus*, family *Rhabdoviridae*. OFV is a bacilliform, nuclear rhabdovirus composed of two segments of single-stranded, negative-sense RNA which infects plants [@Amarasinghe2019; @Dietzgen2014; @Walker2018]. Flat mites from the genus *Brevipalpus* (Trombidiformes: Tenuipalpidae) are the only which have been experimentally shown to transmit dichorhaviruses [@Maeda1998].

@@

Viral infections of suspected leaf samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. All the samples were tested with one step conventional RT-PCR, and were found negative for begomovirus, carlavirus, potyvirus, tospovirus, cucumber mosaic virus and tobacco mosaic virus.

As initial diagnostics were inconclusive, samples were taken of putatively infected plants with ringspot symptoms during July and August of 2020. Leaves were taken from *Liriope* spp. and *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea), which were suspected to be infected, due to both its proximity to infected *Liriope* and the presence of unusually chlorotic leaves (`r citef("fig_2")`). Plant materials were sent to the Florida Department of Agriculture and Consumer Services (FDACS) for identification.

The FDACS determined that the pathogen was OFV using previously published primers and methods to conduct RT-PCR and Sanger sequencing [@Kubo2009a; @Kubo2009b; @RamosGonzalez2015]. Orchid subgroup 1, OFV-Orc, was identified following the methods described in @Kondo2017. Sequencing demonstrated a shared 98\% nucleotide identity with the orchid strain subgroup, OFV-Orc (isolates So and Br with GenBank Accession numbers: AB244418 and MK522807, respectively) [@Kondo2006; @Kondo2017].

These samples from FDACS were subsequently retested by the USDA-ARS, in conjunction with tests of fresh samples from both Alachua and Leon counties. The USDA used RT-PCR, RT-qPCR, and High Throughput Sequencing (HTS) to reconfirm the presence of OFV. RT-PCR and qPCR with Generic R2-Dicho-GF and R2-Dicho-GR primers amplifed ~800 nt of L-gene (RNA2) amplicon [@Roy2020], and OFV-Orc1 and OFV-Orc2 were detected in both *O. intermedius* and *A. elatior* from Leon County. **(insert statement of how much OFV-Orc1 and OFV-Orc2 differ at the nucleotide level)** ***Request from Reviewer 2: deposit and include GenBank accession numbers for the identified virus isolates in Florida? Re:New reports of viruses required that or two other alternative methods for viral detection** **Question from Reviewer 4: Does each isolate reflect a strain?**

@@

After the initial identification by FDACS of OFV-Orc strains, mite samples were collected from symptomatic Asparagaceae in Leon County. Most mites collected were Tenuipalpid mites (flat mites or false spider mites), a pest of ornamental plants, some of which are known to act as vectors for plant viruses [@Childers2003; @Childers2011].

@@

The flat mites collected were initially suspected to belong to *B. californicus* after inspection with phase contrast microscopy. Subsequent observation via DIC microscopy at FDACS agreed with this tentative identification. Unfortunately, the *B. californicus* s.l. species group, *sensu* @Baker1987 is suspected to contain cryptic species [@Childers2011; @Rodrigues2013]. New mite samples were collected from symptomatic liriopogons and *A. elatior* in Leon County and sent to USDA-ARS's Electron and Confocal Microscopy Unit for analysis. Three mite species were recovered and examined under cryo-scanning electron microscopy (Cryo-SEM): *B. californicus* s.l. (`r citef("fig_3")`), *B. obovatus* Donnadieu and *B. confusus* Baker<!--  (`r citef("fig_4")`) -->.

@@

OFV consists of two orchid strains (OFV-Orc1 and OFV-Orc2) and two citrus strains (OFV-Cit1 and OFV-Cit2) [@Roy2020; @Beltran-Beltran2020]. The OFV strains detected in Florida are identical in gene order, content, and genome sequence to the orchid strains of OFV infecting citrus in Hawaii, Mexico, Colombia, and South Africa [@Roy2020; @Beltran-Beltran2020]. Both OFV-Orc1 and OFV-Orc2 infect citrus [@Roy2020], but none of the citrus strains have been reported from any orchid species. The *Brevipalpus* mites collected from liriopogons and *A. elatior* in Leon County were abundant on OFV-infected plants very near to citrus trees, some plants even surrounding the trunk. *B. californicus* s. l. has been reported as a pest of citrus [@Childers2003] and are often collected from citrus fruits [@Vacante2016; @Vacante2010; @Baker1987; @Baker1949]. The proximity of these mite vectors to citrus raises the question: why these trees are not currently infected with OFV-Orc? It is important to note the uncertainty surrounding the vector for OFV-Orc. There are three mite species which have been recovered from OFV-Orc infected plants: *B. obovatus*, and *B. confusus* and *B. californicus* s.l., but only *B. californicus* has been described as a vector of OFV. Even so, the *B. californicus* which we find on liriopogons and *A. elatior* may not be the same species as those found on citrus. Transmission of OFV from populations of *B. californicus* liriopogon/*A. elatior* to citrus may be limited by host preferences, vectorial capacity, viral propagation/circulation in the vector, viral acquisition times, or feeding times required for transmission to citrus. Even so, these types of questions require future study to determine the potential of nolinoidaea to citrus transmission.

@@

In conclusion, detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivated *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and almost eradicated by the mid-1960s [@Childers2003; @Knorr1968a; @Knorr1968b]. An examination of herbarium specimens of Florida citrus found that this historical virus, Citrus leprosis dichorhavirus-N0, is distantly related to the modern strains of OFV [@Hartung2015; @Kitajima2011a; @Roy2020]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the potential threat of different strains of OFV on citrus, which will be a definite concern to the US multi-billion-dollar citrus industry already impacted by the Huanglongbing disease. *B. californicus*, *B. yothersi*, and *B. obovatus* are all present in Florida [@Childers2003; @Akyazi2017] (`r citef("fig_4")`), and are difficult to identify by non-experts, or without advanced methodologies. DNA barcoding [@Armstrong2005] or a similarly simple and accurate method for identification of these mite complexes is vital to identify mite populations which need to be monitored or controlled. By doing so, we can determine the risk OFV-Orc represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

@@

### Table 1: List of Asparagaceae (Nolinoidaea) species with verified cases of orchid fleck virus, collected from the landscape of northern Florida

|Scientific Name|Common Names|County|Strains|
|:-----------|:--------------|:-----|:------|
|*Liriope muscari* cv. 'Gigantea'\* (Decaisne) Bailey | Lilyturf, Orchardgrass, Monkeygrass|Alachua & Leon|OFV-Orc1 & OFV-Orc2|
|*Ophiopogon intermedius*\*\* Don| Aztec Grass, 'Argenteomarginatus'|Leon|OFV-Orc1 & OFV-Orc2|
|*Aspidistra elatior* Blume| Cast Iron Plant, Bar-room Plant|Leon|OFV-Orc1 & OFV-Orc2|
`r tab_nums(name = "tab_1", caption = "* *Liriope muscari* cv. 'Gigantea' has been traditionally classified as *L. gigantea* Hume by @Broussard2007 and @Fantz2015, although this distinction has been challenged by @Wang2014 and @Masiero2020. * * *O. intermedius* is sometimes misclassified as *Liriope muscari* ‘Variegated Evergreen Giant’ [@Fantz2008b, @Fantz2009] or ‘Grandiflora White’ [@Fantz2009]. Both OFV-Orc1 and OFV-Orc2 were detected in all host species.")`

### Figure captions
`r fig_nums(name = "fig_1", caption = "Variety of symptoms expressed by *Liriope* spp. infected with orchid fleck virus: (a) ringspot symptoms on *Liriope muscari* cv. 'Gigantea' (b-c) Details of ringspot symptoms on *L. muscari* cv. 'Gigantea' (d) rust colored spots on *Ophiopogon intermedius*")`

`r fig_nums(name = "fig_2", caption = "Symptoms expressed by *Aspidistra elatior* infected with orchid fleck virus: (a) Detail of leaf chlorosis (b) Chlorosis appears similar to sun damage (c-d) Chlorotic ringspot may indicate early symptoms of OFV")`

@@

<!-- `r fig_nums(name = "fig_4", caption = "Florida is home to other common pest species of *Brevipalpus*, which could be confused with *B. californicus*: (a) *B. phoenicis*, dorsal (b) *B. yothersi*, lateral (c) *B. obovatus*, dorsal.")` -->

@@

<!-- ```{r, fig_4, warning=FALSE} -->
<!-- grid::grid.raster(tiff::readTIFF("figure/other_mites.tif")) -->
<!-- ``` -->
